Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-11 paper/main.tex
adjust interaction section; add morse potential; restructure slightly

% --------------------------------------------------------------------------------------------------
\subsection{Physical Interactions - Mie Potential}

@@

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
By varying its parameters, the Mie potential is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels that correspond to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower number of parameters and more restrictive mathematical formulation
(see \nameref{sec:supplement-morse-potential}).
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low number of parameters that can be estimated.

@@

%###################################################################################################
\paragraph*{Alternative Interaction Potential - Morse Potential}
\label{sec:supplement-morse-potential}
The morse potential~\cite{Morse1929} is a common example of numerically inexpensive interaction
potential.
It was originally designed to describe the interaction of diatomic molecules but has since found use
in other areas~\cite{Breitwieser2021}.
It contains a repulsive and attractive part and is bound from above which makes it particularly
suitable for numerical simulations since unbound potentials can lead to large numerical deviations
thus compromising the prediction of the model.
The potential is of the form
\begin{equation}
    V(r) = V_0\left(1-e^{-\lambda(r-(R_1+R_2))}\right)^2
\end{equation}
and has 3 parameters.
The overall potential strength $V_0$, the radii of the particles $R_1,R_2$ and the potential
stiffness $\lambda$ which controls the rate of change as one travels along the radial axis.
As done with our approach towards the Mie potential, we are free to assign a spatial cutoff $\xi$
to the range of the interaction potential.

>>>>

% --------------------------------------------------------------------------------------------------
\subsection{Physical Interactions}

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
The Mie potential~\cite{Mie1903} is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels corresponding to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower number of parameters and more restrictive mathematical formulation
but may therefore be advantagous in situations of sparse data.
These potentials are a good choice for a simple, interpretable interaction potential with
low number of parameters that can be estimated.

\subsubsection{Mie Potential}

@@

%###################################################################################################
\subsubsection{Morse Potential}
\label{subsec:morse-potential}
The morse potential~\cite{Morse1929} is a common example of numerically inexpensive interaction
potential.
It was originally brought up to describe the interaction of diatomic molecules but has since found
use in other areas~\cite{Breitwieser2021}.
It contains a repulsive and attractive part and is bound from above which makes it particularly
suitable for numerical simulations since unbound potentials can lead to large numerical deviations
thus compromising the prediction of the model.
The potential is of the form
\begin{equation}
    V(r) = V_0\left(1-e^{-\omega(r-(R_1+R_2))}\right)^2
\end{equation}
and has 3 parameters.
The overall potential strength $V_0$, the radii of the particles $R_1,R_2$ and the potential
stiffness $\lambda$ which controls the rate of change as one travels along the radial axis.
We also assign a spatial cutoff $\xi$ to the range of the interaction potential.
Figure~\ref{fig:model-bacterium} (D) shows various configurations of a morse potential.

@@
